The eco-evolutionary impacts of collective straying on metapopulation dynamics

### 2017-11-15 manuscript/FinalDraft3/msv7.tex
fdraft 3

We used the discrete Ricker framework described by Shelton and Mangel \citep{Shelton:2011eq} as the basis for our two-site metapopulation model, with the added effect of the local population $N_i$ mixing with a set proportion $m$ of a remote population $N_j$ that is straying into it.
In this sense, both populations serve simultaneously as donor and recipient populations.
We first assumed that a proportion ${\rm e}^{-Z}$ of both populations survive such that the surviving aggregated population, composed of both local individuals (at site $i$) and incoming strays (from site $j$), is $\left((1-m)N_i(t) + m N_j(t) \right){\rm e}^{-Z}$.
Because local individuals will recruit differently than incoming strays, the recruitment of the aggregate must incorporate two recruitment functions, given by $\left(R_i[\mu_i(t)] (1-m)N_i(t) + R_i[\mu_j(t)] m N_j(t)\right)$.
This mix of individuals is subject to identical compensatory effects, which is determined by the parameter $\beta$.
Taken together, the difference equation that determines changes in population size is

\begin{align}
  &N_i(t+1) = \\ \nonumber
  &\left((1-m)N_i(t) + m N_j(t) \right){\rm e}^{-Z} \\ \nonumber
  &+ \left(R_i[\mu_i(t)] (1-m)N_i(t) + R_i[\mu_j(t)] m N_j(t)\right) \\ \nonumber
  &\times {\rm e}^{-\beta ((1-m)N_i(t) + m N_j(t))},
  \label{eq:N}
\end{align}

\noindent where the equation for $N_j$ mirrors that for $N_i$.

@@

\begin{align}
  &R_i[\mu_i(t)] = \\ \nonumber
  &\int_{-\infty}^\infty r_{\rm max}\exp\left\{\frac{(x_i-\theta_i)^2}{2\tau^2}\right\} {\rm pr}(x_i,\mu_i(t),\sigma^2) {\rm d}x_i +\tilde{P}_i\\ \nonumber
  &= \frac{r_{\rm max} \tau  }{\sqrt{\sigma ^2+\tau ^2}}\exp\left\{-\frac{(\theta_i-\mu_i(t))^2}{2 \left(\sigma ^2+\tau ^2\right)}\right\} +\tilde{P}_i,
  \label{eq:R}
\end{align}

@@

% \noindent{\bf (b) Recruitment over a selective landscape}
Because individuals from the local population are mixed with individuals from the remote population via straying and subsequent reproduction, the resulting trait distribution is a complex mixture of trait distributions (see Supplemental Materials: Appendix I for the derivation, Fig. \ref{fig:PDF}).
We make two simplifying assumptions.
First, we approximate the distribution resulting from the mix of remote and local individuals following reproduction as a Gaussian where $X_i=x_i$ with probability $g(x_i)$.
The expectation of the actual trait distribution as well as the Gaussian approximation are the same, such that ${\rm E}\{g(x_i)\} = w_i\mu_i + (1-w_i)\mu_j$, with weights corresponding to the proportion of the mixed population that are local individuals, $w_i$, and straying individuals, $1-w_i$, where 
\begin{equation}
w_i=\frac{(1-m)N_i(t)}{(1-m) N_i(t) + m N_j(t)}.
\end{equation}
We show that the Gaussian approximation is not only exact with respect to the mean, but also is a good approximation of variance in most relevant cases (see Appendix I in the Supplemental Materials, Fig. \ref{fig:PDF}).
Thus, strays can successfully reproduce and introduce their genotypes into the recipient population, which is supported by observations in wild populations \citep{Jasper:2013cc}.
Secondly, we assumed that changes in trait variance through time are minimal, such that $\sigma^2$ is constant over time, which is a common simplification in eco-evolutionary models of population dynamics \citep{Lande:1976ga,Schreiber:2011wx,Gilbert:2014ee,Gibert:2015kc}.

@@

\begin{align}
  \label{eq:mu}
  \mu_i(t+1) &= w_i\mu_i(t) + (1-w_i)\mu_j(t) \\ \nonumber
  &+ h^2\sigma^2\frac{\partial}{\partial \mu_i}\ln\left(w_i R_i[\mu_i(t)] + (1-w_i)R_i[\mu_j(t)]  \right),
\end{align}

>>>>

We used the discrete Ricker framework described by Shelton and Mangel \citep{Shelton:2011eq} as the basis for our two-site metapopulation model, with the added effect of the local population $N_i$ mixing with a set proportion $m$ of a remote population $N_j$ that is straying into it.
In this sense, both populations serve simultaneously as donor and recipient populations.
We first assumed that a proportion ${\rm e}^{-Z}$ of both populations survive such that the surviving aggregated population, composed of both local individuals (at site $i$) and incoming strays (from site $j$), is $\left((1-m)N_i(t) + m N_j(t) \right){\rm e}^{-Z}$.
Because local individuals will recruit differently than incoming strays, the recruitment of the aggregate must incorporate two recruitment functions, given by $R_i[w_i\mu_i(t) + (1-w_i)\mu_j(t)]\left((1-m)N_i + mN_j\right)$ (Eq. \ref{eq:R}).
This mix of individuals is subject to identical compensatory effects, which is determined by the parameter $\beta$.
Taken together, the difference equation that determines changes in population size is

\begin{align}
  &N_i(t+1) = \\ \nonumber
  &\left((1-m)N_i(t) + m N_j(t) \right){\rm e}^{-Z} \\ \nonumber
  &+ R_i[w_i\mu_i(t) + (1-w_i)\mu_j(t)]\left((1-m)N_i + mN_j\right) \\ \nonumber
  &\times {\rm e}^{-\beta ((1-m)N_i(t) + m N_j(t))},
  \label{eq:N}
\end{align}

\noindent where

\begin{equation}
w_i=\frac{(1-m)N_i(t)}{(1-m) N_i(t) + m N_j(t)},
\end{equation} 
and the equation for $N_j$ mirrors that for $N_i$.

@@

\begin{align}
  &R_i[w_i\mu_i(t) + (1-w_i)\mu_j(t)] = \\ \nonumber
  &\int_{-\infty}^\infty r_{\rm max}\exp\left\{\frac{(x_i-\theta_i)^2}{2\tau^2}\right\} {\rm pr}(x_i,w_i\mu_i(t) + (1-w_i)\mu_j(t),\sigma^2) {\rm d}x_i +\tilde{P}_i\\ \nonumber
  &= \frac{r_{\rm max} \tau  }{\sqrt{\sigma ^2+\tau ^2}}\exp\left\{-\frac{(\theta_i-(w_i\mu_i(t) + (1-w_i)\mu_j(t)))^2}{2 \left(\sigma ^2+\tau ^2\right)}\right\} +\tilde{P}_i,
\end{align}

@@

% \noindent{\bf (b) Recruitment over a selective landscape}
Because individuals from the local population are mixed with individuals from the remote population via straying and subsequent reproduction, the resulting trait distribution is a complex mixture of trait distributions (see Supplemental Materials: Appendix I for the derivation, Fig. \ref{fig:PDF}).
We make two simplifying assumptions.
First, we approximate the distribution resulting from the mix of remote and local individuals following reproduction as a Gaussian where $X_i=x_i$ with probability $g(x_i)$.
The expectation of the actual trait distribution as well as the Gaussian approximation are the same, such that ${\rm E}\{g(x_i)\} = w_i\mu_i + (1-w_i)\mu_j$, with weights corresponding to the proportion of the mixed population that are local individuals, $w_i$, and straying individuals, $1-w_i$.
We show that the Gaussian approximation is not only exact with respect to the mean, but also is a good approximation of variance in most relevant cases (see Appendix I in the Supplemental Materials, Fig. \ref{fig:PDF}).
Thus, strays can successfully reproduce and introduce their genotypes into the recipient population, which is supported by observations in wild populations \citep{Jasper:2013cc}.
Secondly, we assumed that changes in trait variance through time are minimal, such that $\sigma^2$ is constant over time, which is a common simplification in eco-evolutionary models of population dynamics \citep{Lande:1976ga,Schreiber:2011wx,Gilbert:2014ee,Gibert:2015kc}.

@@

\begin{align}
  \label{eq:mu}
  \mu_i(t+1) &= w_i\mu_i(t) + (1-w_i)\mu_j(t) \\ \nonumber
  &+ h^2\sigma^2\frac{\partial}{\partial \mu_i}\ln\left(R_i[w_i \mu_i(t)+ (1-w_i)\mu_j(t)] \right),
\end{align}

### 2017-11-15 manuscript/FinalDraft3/supp.tex
fdraft 3

>>>>

\section*{I. Gaussian approximation for the mixed trait distribution}
We assume a Gaussian distribution to approximate the more complex trait distribution that results when two populations with two distinct trait distributions are linked by gene flow and reproduction.
Here and throughout, we refer to the resident population as $i$ and the straying population as $j$, though both populations stray simultaneously.
The Gaussian approximation is assumed to have a mean that is equal to the mean of the mixed trait distribution $(1-m)\mu_i + m\mu_j$ with a constant variance $\sigma^2$, equal to the original trait variance prior to mixing.
We employ this approximation because a mixed trait distribution that remixes and recombines (through reproduction) at each time-step becomes complicated very quickly, as we shall demonstrate.
\emph{Importantly, because the evolutionary framework tracks only the mean and not the variability through time, this approximation is exact with respect to the first moment.}
Accordingly, the only attribute that is approximate is the assumption of normality.
Below we will show that a constant variance of $\sigma^2$ for the mixed trait distribution is a good approximation in most relevant cases, and becomes increasingly accurate when 1) generations become less overlapping (i.e. the case of semelparity), and 2) with density-dependent straying ratios.

{\bf The trait distribution at $\bm{t=0}$.} At time $t=0$, it is assumed that individuals from two separate populations have not yet mixed. 
Thus, the mean trait values for both populations are equal to the optimal trait values that accord to their respective habitats $\mu_i(t=0) = \theta_i$, and that their trait values are distributed as $X_i=x_i \sim g(x_i,\mu_i,\sigma) = \mathcal{N}(\mu_i,\sigma)$.
Here and henceforth we use upper-case to denote random variables, and lower-case to denote specific values of random variables.

{\bf The trait distribution at $\bm{t=1}$.} At time $t=1$, dispersal of individuals from site $j$ to site $i$, and vice versa, has begun.
The proportion of individuals that disperse to and from each population is given by the stray ratio $m$.
The calculation of the mixed distribution that results from dispersal and subsequent recruitment incorporates the following processes:
\emph{1)} dispersal of strays into the local population from the remote population, and vice versa;
\emph{2)} recruitment of juveniles that are produced from this mixed distribution (under the assumption of random mating); 
\emph{3)} mortality of individuals in the adult class.
These events occur in the order in which they are listed.

The mixed trait distribution for site $i$ immediately following dispersal is simply a mixed Gaussian with weights determined by the proportion of the population that are residents $i$ and the proportion that are strays $j$, such that 
\begin{equation}
  p(x_{\rm disperse}) = w_i g(x_i,\mu_i,\sigma) + (1-w_i)g(x_j,\mu_j,\sigma),
\end{equation}
where the proportion of the population $i$ composed of \emph{resident} individuals is
\begin{equation}
  w_i = \frac{(1-m)N_i}{(1-m)N_i + mN_j},
\end{equation}
and the proportion of population $i$ composed of strays is $(1-w_i)$.
Once individuals in both habitats mix, they mate without preference.
Recruits are assumed to have trait values intermediate to their parents such that $x_{\rm r} = (x_{1} + x_{2})/2$ where trait values $X_1=x_{1}\sim p(x_{\rm disperse})$ and $X_2=x_{2}\sim p(x_{\rm disperse})$.
We thus assume that parental traits are drawn independently from the mixed distribution $p(x_{\rm disperse})$, conditional on the requirement that $x_2 = 2x_r - x_1$, which we get when rearranging the equation for $x_{\rm r}$.
Incorporating this condition and that traits are independently drawn from the mix of residents and strays yields the probability density for the trait value of recruits,
\begin{equation}
  q(x_{\rm r}) = \int_{-\infty}^\infty 2p(x_1)p(2x_{\rm r}-x_1){\rm d}x_1,
\end{equation}
where the factor of 2 comes from...
The mean of the recruit trait distribution is calculated to be
\begin{equation}
{\rm E}\{q(x_{\rm r})\}=(1-m)\mu_i + m\mu_j,
\end{equation} 
and the variance is
\begin{equation}
  {\rm Var}\{q(x_{\rm r})\} = \frac{m^2 N_j^2 \left(4 \mu_j^2+\sigma ^2\right)+(m-1)^2 N_i^2 \left(4 \mu_i^2+\sigma ^2\right)-(m-1) m N_i N_j \left(\mu_i^2+6 \mu_i \mu_j+\mu_j^2+2 \sigma ^2\right)}{2 (-m N_i+m N_j+N_i)^2}.
\end{equation}

The relative contribution of trait values from resident parents, stray parents, and recruits depends on the numbers of each within the mixed population.
The number if residents is $(1-m)N_i$, and the number of strays is $mN_j$.
However, because adult mortality occurs after reproduction, we must adjust these numbers to account for only those who survive, which is given by ${\rm e}^{-Z}$.
As $Z\rightarrow \infty$, the population becomes semelparous, as the adult class gives way to new recruits at each timestep. 
Accordingly, the number of adult residents and strays is given by $(1-m)N_i{\rm e}^{-Z}$ and  $mN_i{\rm e}^{-Z}$, respectively.
In contrast, the number of recruits is given by the combined contribution of resident and straying individuals to recruit production as $N_r = ((1-m)N_1 R[\mu_i] + mN_2 R[\mu_j]){\rm e}^{(-\beta((1-m)N_i + mN_j))}$ where $R$ is the recruitment rate, which differs for resident vs. straying individuals (see manuscript, Eq 2).

\begin{figure}
  \captionsetup{justification=raggedright,
singlelinecheck=false
}
\centering
\includegraphics[width=0.7\textwidth]{fig_PDF.pdf}
\caption{
Clockwise from the upper-left. The calculated probability distribution function (PDF) for individuals in population $i$ after immigration from population $j$, recruitment, and mortality (blue) vs. the Gaussian approximation used for the dynamical system (green), with
a) low stray ratio, large straying population, overlapping generations: $m=0.1$, $Z=0.5$;
b) high stray ratio, overlapping generations: $m=0.5$, $Z=0.5$;
c) high stray ratio, non-overlapping generations: $m=0.5$, $Z=1000$;
d) density-dependent stray ratio, non-overlapping generations: $m=m_0(1-N_i/(C+N_i))$, $Z=1000$, where $m_0=0.2$ and $C=1000$.
The mean and variability of both distributions is shown by a point (mean) and line ($\sigma$) at the base of each distribution.
} \label{fig:PDF}
\end{figure}

The proportion of individuals from the resident, straying, and recruitment classes following mortality at site $i$ are then $c_i$, $c_j$, and $c_r$ respectively, where
\begin{align}
c_i &= \frac{(1-m)N_i{\rm e}^{-Z}}{(1-m)N_i{\rm e}^{-Z} + m N_j{\rm e}^{-Z} + N_r} \\ \nonumber
c_j &= \frac{m N_j{\rm e}^{-Z}}{(1-m)N_i{\rm e}^{-Z} + m N_j{\rm e}^{-Z} + N_r} \\ \nonumber
c_r &= \frac{N_r}{(1-m)N_i{\rm e}^{-Z} + m N_j{\rm e}^{-Z} + N_r}.
\end{align}
where $c_i + c_j + c_r= 1$.
The trait value distribution for time $t=1$ is then calculated as a mix of trait distributions from 1) surviving residents, 2) surviving strays, and 3) their offspring, such that
\begin{equation}
  g_{{\rm next}}(x_i) = c_i g(x_i) + c_j g(x_j) + c_r q(x_r).
\end{equation}
The expectation of this distribution is
\begin{equation}
{\rm E}\{g_{\rm next}(x_i)\}=(1-m)\mu_i + m\mu_j,
\end{equation} and the variance is
% variance a that has a closed form solution, but that is not easily readable.
\begin{align}
  {\rm Var}\{g_{{\rm next}}(x_i)\}=&\frac{(\mu_i-\mu_j)^2 (2 (m-1)^2 N_i^2 c_j^2-(m-1) N_i c_j (2 (m-1) N_i+m N_j (1-4 c_i))}{2 (-m N_i+m N_j+N_i)^2} \\ \nonumber
  &+ \frac{m N_j (c_i-1) (-m N_i+2 m N_j c_i+N_i))-\sigma ^2 (c_i+c_j+1) (-m N_i+m N_j+N_i)^2}{2 (-m N_i+m N_j+N_i)^2}.
\end{align}

Importantly, ${\rm E}\{g_{{\rm next}}(x_i)\}$ is exactly equal to the trait mean for the original mix, as well as the trait mean for recruits.
Thus, a Gaussian approximation distributed as $\mathcal{N}(w_i\mu_i + (1-w_i)\mu_j,\sigma)$ will not alter the dynamics under the assumption of constant variance in Eq 4.
Moreover, the mixed distribution described above becomes more complex at the next timestep, as two multi-modal trait distributions then mix and recombine through recruitment.
Although the complexity of the true distribution is increased and becomes quickly intractable, one expects the actual form to eventually approximate a Gaussian based on the central limit theorem.
One could make the argument that a Gaussian approximation does not represent the multi-modality in trait values inherent to a mixed distribution, and thus underestimates variance.
Although the Gaussian approximation exactly captures the expectation, we next explore to what extent the approximation does and does not capture the variability expressed in $g(x_{\rm next})$.

%Discussion of factors that increase or decrease accuracy of the approximation
The Gaussian approximation $\mathcal{N}(w_i\mu_i + (1-w_i)\mu_j,\sigma)$ will generally underestimate the true variance of the mix, but not by much.
If the straying ratio is low, (ca. 0.0 to 0.2), the Gaussian approximation tends to be very similar to the actual distribution (Fig. \ref{fig:PDF}a).
As the straying ratio increases, so does the bimodality of the actual distribution, and this magnifies our underestimate of variance (Fig. \ref{fig:PDF}b).
The interesting dynamics observed in our model generally occur between $m=0.05$ and $m=0.2$, where estimates of variance are similar.
Decreasing the overlap between generations also leads to greater similarity in the actual distribution and the Gaussian approximation, and the potential bias introduced by bimodality will be less for cases of non-overlapping generations, or semelparity (Fig. \ref{fig:PDF}c).
Most importantly, density-dependent straying increases the accuracy of our Gaussian approximation by suppressing the straying ratio (Fig. \ref{fig:PDF}d).

Because the Gaussian approximation is exact to first order, and generally provides a good approximation of realized variance, we suggest that the use of Lande's formulation for the change in mean trait values through time under the assumption of constant variance is appropriate (see Eq. 4 in main text).
Because our estimate of variance is generally an underestimate, the rate of recruitment will be a slight overestimate (Eq. 2, main text), and the timescale of trait evolution will generally be slower (Eq. 4, main text).
Changes in the rate of recruitment (Fig. \ref{fig:symmetry}) and heritability do not have an impact on our qualitative findings, and this suggests any bias that our approximation introduces by underestimating true variance is minimal.

\section*{II. Additional Figures}

% \begin{figure}
%   \captionsetup{justification=raggedright,
% singlelinecheck=false
% }
% \centering
% \includegraphics[width=0.7\textwidth]{fig_contSD.pdf}
% \caption{
% Magnitude of difference in the standard deviation for the calculated PDF for individuals in population $i$ after 1) immigration from population $j$, 2) recruitment, and 3) adult mortality vs. the Gaussian approximation where $\sigma=1$.
% Values $<1$ reflect underestimates of variability; values $>1$ reflect overestimates.
% Top row is for constant straying ratio, and the bottom row is for density-dependent straying ratio under different treatments of habitat heterogeneity (left column $\Delta \theta=5$; right column $(\Delta \theta=8$).
% } \label{fig:contourSD}
% \end{figure}
%

### 2017-11-15 manuscript/oldversions/msv2.tex
fdraft 3

We have shown that dispersal between populations, coupled with localized selection against donor phenotypes, has a large and nonlinear impact on dynamic properties that directly impacts metapopulation robustness, and by extension persistence.
Although there are various ways to measure `robustness' (REF:TREE), we employ two measures: 
1) the average-CV portfolio effect, a statistical metric commonly used to assess the fragility of salmon populations; 
2) the recovery time, defined here as the time required for the aggregate metapopulation biomass $N_T$ to return to its steady state after different types of induced disturbances, which is mechanistically linked to the likelihood of persistence (REF).
In our minimal model of dispersal and natural selection between two populations, we show that these phenomenological and mechanistic descriptors of metapopulation robustness are tightly coupled (figure \ref{fig:PE}d).

%Alternative steady states & critical slowing
The immigration of dispersing strays with trait values far from the local optimum generally lowers the combined steady state biomass $N_T$ and increases the likelihood that the system gives way to alternative steady states, pushing one of the populations close to extinction.
The formation of alternative stable states in our system is a relatively common example of spatial pattern formation, and occurs above a threshold straying rate.
Pattern formation can occur as a consequence of myriad ecological processes, including habitat selection (REFS), aggregation of individuals (REFS), local environmental conditions (REFS), or interspecific interactions like competition (REFS). 
It is also well known to occur with the onset of spatially or diffusion-induced instabilities (Turing instabilities, REFS). 
Here, we describe an interesting alternative mechanism through which pattern formation in local densities can occur, when the onset of movement mediated maladaptation leads to local differences in reproductive rates that yield differences in densities that would, in the absence of straying, be the same. 
This mechanism of pattern formation through the simultaneous interplay between ecological and evolutionary dynamics may open up new areas of research.

To what extent the existence of alternative stable states promotes or diminishes the system as a whole is a matter of perspective.
On the one hand, one population is pushed to a low-density state (figure \ref{fig:traj}) with a lower total biomass of the aggregate (figure ]ref{fig:PE}a), suggesting that the system is more fragile.
On the other hand, the presence of \emph{just enough} straying to cause formation of alternative stable states both increases the portfolio effect (figure \ref{fig:PE}c) and reduces the rate of return -- particularly if straying is density dependent (figure \ref{fig:relax}c,d) as might be common in systems with collective navigation (REF).

>>>>

We have shown that dispersal between populations, coupled with localized selection against donor phenotypes, has a large and nonlinear impact on dynamic properties that directly impacts metapopulation robustness, and by extension persistence.
Although there are various ways to measure `robustness' (REF:TREE), we employ two measures: 
1) the average-CV portfolio effect, a statistical metric commonly used to assess the fragility of populations; 
2) the recovery time, defined here as the time required for the aggregate metapopulation biomass $N_T$ to return to its steady state after different types of induced disturbances, which is mechanistically linked to the likelihood of persistence (REF).
In our minimal model of dispersal and natural selection between two populations, we show that these phenomenological and mechanistic descriptors of metapopulation robustness are tightly coupled (figure \ref{fig:PE}d).

%Alternative steady states & critical slowing
The immigration of dispersing individuals with trait values far from the local optimum generally lowers the combined steady state biomass $N_T$ and increases the likelihood that the system gives way to alternative steady states, pushing one of the populations close to extinction.
The formation of alternative stable states in our system is a relatively common example of spatial pattern formation, and occurs above a threshold straying rate.
Pattern formation can occur as a consequence of myriad ecological processes, including habitat selection (REFS), aggregation of individuals (REFS), local environmental conditions (REFS), or interspecific interactions like competition (REFS). 
It is also well known to occur with the onset of spatially or diffusion-induced instabilities (Turing instabilities, REFS). 
Here, we describe an interesting alternative mechanism through which pattern formation in local densities can occur, when the onset of movement mediated maladaptation leads to local differences in reproductive rates that yield differences in densities that would, in the absence of straying, be the same. 
%This mechanism of pattern formation through the simultaneous interplay between ecological and evolutionary dynamics may open up new areas of research.

To what extent the existence of alternative stable states promotes or diminishes the system as a whole is a matter of perspective.
On the one hand, one population is pushed to a low-density state (figure \ref{fig:traj}) with a lower total biomass of the aggregate (figure ]ref{fig:PE}a), suggesting that the system is more fragile.
On the other hand, the presence of \emph{just enough} straying to cause formation of alternative stable states both increases the portfolio effect (figure \ref{fig:PE}c) and reduces the rate of return -- particularly if straying is density dependent (figure \ref{fig:relax}c,d) as might be common in systems with collective navigation (REF).
Importantly, the asymmetry in population densities also leads an asymmetry in the mean trait values between sites: because the high-density population contains the majority of biomass, the proportion of the low-density population that is composed of incoming strays is generally large, and this leads to mean trait values in both sites skewed towards that of the high-density site.
This has important ramifications for metapopulation robustness, as measured by both the portfolio effect as well as recovery time.
